\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}

\begin{document}

\begin{center}
{\Large A Treatment-Effect-Informed Learned Environment for Reinforcement Learning-Based Farmland Consolidation Planning}
\end{center}

\section*{Author Information}

Ning Zhou$^{a}$ and Xiang Jing$^{a,*}$

\noindent $^{a}$School of Software and Microelectronics, Peking University, Beijing 100871, China

\noindent $^{*}$Corresponding author: Xiang Jing, \href{mailto:jingxiang@pku.edu.cn}{jingxiang@pku.edu.cn}

\section*{Article Type}

Research article.

\section*{Funding}

No funding statement was present in the source manuscript. The authors should confirm the final funding statement before submission.

\section*{Declaration of Interest}

The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

\section*{CRediT Author Statement}

Ning Zhou: Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data curation, Writing - original draft, Visualization.

Xiang Jing: Conceptualization, Methodology, Supervision, Writing - review and editing, Project administration.

\section*{Data and Code Availability}

The manuscript uses administrative cadastral and land-use data for Bishan District, Chongqing. Raw administrative data may be subject to access restrictions. Derived result summaries and analysis scripts can be provided for peer review in anonymized form, and a public repository can be prepared after removing restricted data and author-identifying metadata.

\section*{Generative AI Statement}

Generative AI tools were used for language polishing and submission-package preparation. The authors reviewed and edited the output and take full responsibility for the content of the submitted manuscript.

\end{document}
